The Unicode Cookbook \linebreak for Linguists

### 2018-06-05 book/chapters/implementation.tex
work towards paperhivers chapter 8

Note that our two libraries have rather different implementation histories, thus they may not give the same results in all situations (as discussed in Chapter~\ref{orthography-profiles}). However, we do provide extensive test suites for each implementation that follow standard practices to make sure that results are correct. Users should refer to these tests and to the documentation in each release for specifics about each implementation. Note that due to the different naming convention practices in Python and R, function names differ between the two libraries. Also, the performance with larger datasets may not be comparable between the Python and R implementations. In sum, our two libraries should be considered as proof of concept and not as the final word on the practical application of the specifications discussed in the previous chapter. In our experience, the current versions are sufficiently fast and stable to be useful for academic practice (e.g.\ checking data consistency, or analyzing and transliterating small to medium sized data sets), but they should probably not be used for full-scale industry applications without adaptation.

@@

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple to install (via the Terminal App) free and open source package management system. Follow the instructions on the the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

@@

To install the Python package from the Python Package Index (PyPI) run:

@@

You can also install the \texttt{segments} package with from the GitHub repository:\footnote{\url{https://github.com/cldf/segments}}

@@

\noindent We can also load an orthography profile and tokenize an input string with it. In the data directory we've placed an example orthography profile. Let's have a look at it using \texttt{more} on the command line.

@@

\noindent An orthography profile is delimited UTF-8 text file (here we use tab as a delimiter for reading ease). The first column must be labelled \texttt{Grapheme}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns IPA and XSAMPA, which are mappings from our graphemes to their IPA and XSAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ''quote that     string''!

@@

\noindent This example shows how can we tokenize input text into our orthographic specification. We can also segment graphemes and transliterate them into other formats, which is useful when you have sources with different orthographies, but you want to be able to compare them using a single representation like IPA.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note the in large text the frequency column may signal errors in the input, such as typos, if they occur with a very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \texttt{Grapheme} column and then add additional columns for transliterations or comments.

@@

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called vignette with
explanations and examples.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) shiny webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R.

@@

These executables can be 
used from here, or they can be linked and/or copied to any location as wanted. 
For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. The webapps are 
included inside the \texttt{qlcData} package and can be started with the 
helper function (in R): \texttt{launch\_shiny('tokenize')}.

@@

The first example of how to use these functions concerns finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just an surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using here the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Chapter~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single code points, others are combinations
of two code points (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode code points,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC normalization turns everything into the
precomposed characters, while NFD normalization separates everything into base
characters with combining diacritics. Splitting by code points (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

<<>>=
# after NFC normalization unicode code points have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives yet another structure of the code points
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

@@

It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

<<tidy = FALSE>>=
# note that NFC and NFD normalization are identical
# for unicode grapheme definitions
profile_5 <- write.profile(example, normalize = "NFD")
profile_6 <- write.profile(example, normalize = "NFC")
@

@@

Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see above for how to install the bash executable). 
Exactly this example is also included in the help file of the executable.

@@

Not yet everything is correct in this graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behavior. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

@@

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. the have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile, because most
hand-written specifications of graphemes will first write the general rule,
followed by more specific regularities. To get the right tokenization, these 
rules should in most cases be applied in reverse order.

Note that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied because a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specified order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules is rather easy to predict.

@@

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurrence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard-coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
For example, we are adding an extra line to the profile from above, resulting in
the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
graphemes are based on regular expressions so you can also use regular
expressions in the description of the context. For such contextually specified
graphemes to be included in the graphemic parsing we have to specify the option
\texttt{regex = TRUE}. This contextually specified grapheme should actually be
handled first, so we could try \texttt{ordering = NULL}. However, we can also
explicitly specify that rules with contextual information should be applied
first by using \texttt{ordering = "context"}. That gives the right results for
this toy example.

@@

To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
that we have changed the name of the transliteration-column, so we have to tell
the tokenization process to use this column to transliterate. This is done by
adding the option \texttt{transliterate = "IPA"}.

@@

Another equivalent possibility is to use a column \texttt{Class} to specify a
class of graphemes, and then use this class in the specification of context.
This is useful to keep track of recurrent classes in larger profiles. You are
free to use any class-name you like, as long as it doesn't clash with the rest
of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
give the same result as obtained previously by using a regular expression.

>>>>

Note that our two libraries have rather different implementation histories, thus they may not give the same results in all situations (as discussed in Chapter~\ref{orthography-profiles}). However, we do provide extensive test suites for each implementation that follow standard practices to make sure that results are correct. Users should refer to these tests and to the documentation in each release for specifics about each implementation. Note that due to the different naming convention practices in Python and R, function names differ between the two libraries. Also, the performance with larger datasets may not be comparable between the Python and R implementations. In sum, our two libraries should be considered as proofs of concept and not as the final word on the practical application of the specifications discussed in the previous chapter. In our experience, the current versions are sufficiently fast and stable to be useful for academic practice (e.g.\ checking data consistency, or analyzing and transliterating small to medium sized data sets), but they should probably not be used for full-scale industry applications without adaptation.

@@

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note that there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple-to-install (via the Terminal App) free and open source package management system. Follow the instructions on the the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

@@

To install the Python package \texttt{segments} from the Python Package Index (PyPI) run:

@@

You can also install the \texttt{segments} package from the GitHub repository:\footnote{\url{https://github.com/cldf/segments}}

@@

\noindent We can also load an orthography profile and tokenize an input string with it. In the data directory, we've placed an example orthography profile. Let's have a look at it using \texttt{more} on the command line.

@@

\noindent An orthography profile is a delimited UTF-8 text file (here we use tab as a delimiter for reading ease). The first column must be labelled \texttt{Grapheme}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns IPA and XSAMPA, which are mappings from our graphemes to their IPA and XSAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ``quote that     string''!

@@

\noindent This example shows how we can tokenize input text into our orthographic specification. We can also segment graphemes and transliterate them into other formats, which is useful when you have sources with different orthographies, but you want to be able to compare them using a single representation like IPA.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note that, in large texts, the frequency column may signal errors in the input, such as typos, if a grapheme occurs with very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \texttt{Grapheme} column and then add additional columns for transliterations or comments.

@@

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. The package includes
help files with illustrative examples, and also a so-called vignette with
explanations and examples.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) \texttt{Shiny} webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-like machine. The exact location of these
executables is best found after installation of R the packages. The
location can be found by the following command in R.

@@

These executables can be 
used in the resulting file path, or they can be linked and/or copied to any location as wanted. 
For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{Shiny} framework for the R functions. The webapps are 
included inside the \texttt{qlcData} package and can be started with the 
helper function (in R): \texttt{launch\_shiny('tokenize')}.

@@

The first example of how to use these functions concerns finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just a surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters (called user-perceived characters; see Chapter~\ref{the-unicode-approach} for an explanation). The results are shown 
in Table \ref {tab:profile1}. As it 
turns out, some of these graphemes are single code points, others are combinations
of two code points (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode code points,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions. The result is 
shown in Table \ref{tab:profile2}.

@@

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC normalization turns everything into the
precomposed characters, while NFD normalization separates everything into base
characters with combining diacritics. Splitting by code points (i.e. \texttt{sep~=~""}) 
shows the results of these two normalizations in Tables \ref{tab:profile3} \& \ref{tab:profile4}.

<<>>=
# after NFC normalization Unicode code points have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives another structure of the code points
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

@@

It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFD or NFC, as shown in Tables \ref{tab:profile5} \& \ref{tab:profile6}, 
by using the default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly below.

<<tidy = FALSE>>=
# note that NFC and NFD normalization are identical
# for Unicode grapheme definitions
profile_5 <- write.profile(example, normalize = "NFD")
profile_6 <- write.profile(example, normalize = "NFC")
@

@@

These different profiles can also be produced using the bash
executable \texttt{writeprofile} (see above for how to install the bash executable). 
This example is also included in the help file of the executable.

@@

Everything is not yet correct with the graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behavior. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

@@

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile, because most
hand-written specifications of graphemes will first write the general rule,
followed by more specific regularities. To get the right tokenization, these 
rules should in most cases be applied in reverse order.

Note that different ordering of the rules does not result in 
feeding and bleeding effects found with finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a following rule from applying. Feeding is the opposite: a
specific rule will only be applied because a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here is 
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specified order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the applied rules is rather easy to predict.

@@

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurrence in the data. The lines
         in the file are re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the grapheme is to be
separated.\footnote{The column names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard-coded, so these exact column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
For example, we are adding an extra line to the profile from above, resulting in
the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
graphemes are based on regular expressions so you can also use regular
expressions in the description of the context. For such contextually specified
graphemes to be included in the graphemic parsing we have to specify the option
\texttt{regex = TRUE}. This contextually specified grapheme should actually be
handled first, so we could try \texttt{ordering = NULL}. However, we can also
explicitly specify that rules with contextual information should be applied
first by using \texttt{ordering = "context"}. That gives the right results for
this toy example, as shown in Table \ref{tab:profile_skeleton2}.

@@

To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
that we have changed the name of the transliteration column, so we have to tell
the tokenization process to use this column to transliterate. This is done by
adding the option \texttt{transliterate = "IPA"}.

@@

Another equivalent possibility is to use a column \texttt{Class} to specify a
class of graphemes, and then use this class in the specification of context.
This is useful to keep track of recurrent classes in larger profiles. You are
free to use any class-name you like, as long as it does not clash with the rest
of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
give the same result as obtained previously by using a regular expression.

### 2018-06-05 book/chapters/ipa_meets_unicode.tex
work towards paperhivers chapter 8

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different unicode code points that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
Each of these tables shows a typical glyph, and then lists the Unicode Code point,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@

>>>>

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different Unicode code points that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
Each of these tables shows a typical glyph, and then lists the Unicode Code point,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@
